---
title: "An Introduction to finalfit"
---

### Introduction

For Project A, The student utilized the finalfit package to create clean tables for their logistic regression model. finalfit is an open-source package to be used with R Studio to create clean, publication-ready tables for the data analysis model of your choosing. The package compiles your numbers into a table, grouping data in an efficient, automated way to present in a paper or poster. finalfit is a great programming package for beginners in data analytics research.

### Use Cases and Limitations

:::::: columns
::: {.column width="45%"}
#### finalfit Can:

- Summarize quantitative variable values by a categorical variable
- Summarize regression model results in table format
- Summarize regression model results into a plot
:::

::: {.column width="10%"}
<!-- This acts as empty spacer columns -->
:::

::: {.column width="45%"}
#### finalfit Cannot:

- Analyze results from advanced machine learning models
- Handle large amounts of levels for categorical variables
- Perform Bayesian modeling
:::
::::::

Generally speaking, finalfit is most useful when \[insert here\]

### Example Utilization

Let's look at one example situation where finalfit can make data analysis much easier than it would be in base R. Let's say we have the following question:

> "Using the `infert` dataset, which tracks infertility after spontaneous and induced abortion, what factors strongly indicate whether a patient experienced secondary infertility?"

Finalfit is the perfect tool for a question like this. We can define `case`, which indicates whether a patient experienced secondary infertility (`1`) or served as a control (`0`), as our dependent variable. We can then select factors like `age` (in years), `parity` (count of total pregnancies), `induced` (number of prior induced abortions), and `spontaneous` (number of prior spontaneous abortions) as our explanatory variables.

The R cells below demonstrate the process of implementing this vision:

```{r}
#| warning: false

# Import Libraries
library(finalfit)
library(dplyr)

# Check column types
str(infert)
```

One thing to note: we would like `case`, `induced`, and `spontaneous` to be treated by finalfit as categorical variables, but R recognizes them as numeric integers. We must correct that using `as.factor()` before we build our tables, which is made easy using dplyr's `mutate()` function. ([What's dplyr?](dplyr-guide.qmd))

```{r}
#| warning: false

# Prepare data by casting categorical variables to factors
df <- infert |> 
  mutate(
    case = factor(case, levels = c(0, 1), labels = c("Control", "Case")),
    across(c(induced, spontaneous), as.factor)
  )

# Define dependent and explanatory variables
dependent <- "case"
explanatory <- c("age", "parity", "induced", "spontaneous")

# Generate the full regression model table in one function call
ff_table <- df |>
  finalfit(dependent, explanatory)

# Print clean output
knitr::kable(ff_table, row.names = FALSE)
```

This table unlocks a litany of insights on its own, such as:

* With all other variables controlled, having 1 or 2 induced abortions increases a patient's odds of secondary infertility by factors of 3.12 and 11.09 respectively, both with statistically significant *p* values.
* Parity becomes a protective variable when adjusting for prior abortions, decreasing a woman's odds of secondary infertility by approximately 51% for each additional pregnancy they've had.
* Age has no significant impact on secondary infertility, as its univariate and multivariate *p* values are 0.956 and 0.084 respectively, both above the maximum threshold of 0.05 for a relationship to be statistically significant.

This is just one example of how finalfit can help data scientists like yourself. The package also offers a variety of modeling tools, so consider utilizing it if you find yourself wanting to analyze predictors of a variable's outcome during your research.

### Learn More

If you would like to learn more about finalfit, visit [its website](https://finalfit.org/index.html) for more information. 