---
title: "Modeling Data"
---

### Cleaning Data

### Choosing a Model

### Linear Modeling (lm, mlm, glm)

### Time-to-Event Modeling (survival)

Time-to-event modeling, also known as survival analysis, is effective at adding a temporal element to predictive modeling. Whereas other binary output models are structured to predict the likelihood of a certain event happening, time-to-event modeling predicts the likelihood of an event happening at each passing unit of time. Survival modeling is also equipped to handle the issue of censored data. Many datasets where survival modeling would be effective only have two possible outcomes: the event in question eventually happens, or the observation is eventually cut off (censored) and the final outcome is unknown. For censored outcomes, the model will incorporate the information that is available to it (the length of time wherein an event did not occur) when developing its probability curve.

When using time-to-event modeling in R, it is most common to use the `survival` package. The package includes an example dataset, `lung`, which documents survival data from over 200 advanced lung cancer patients. Let's create a survival model based on `lung` to estimate a patient's likelihood of survival at each passing unit of time.

By using the command `?lung` in R, we can learn the following information about our dataset:

* Patient outcome is denoted as `status`, with 1 = censored and 2 = dead.
* `time` denotes the length of time in days that the patient survived before dying or being censored.
* `ph.ecog` denotes the patient's ECOG performance score, with 0 being the best (asymptomatic), and 4 being the worst (bed-bound).
* `ph.karno` and `pat.karno` denote the physician and patient's subjective performance scores for the patient's status respectively on a scale of 1-100.
* `meal.cal` denotes the average number of calories consumed per meal during treatment and `wt.loss` denotes the patient's weight loss in pounds in the last six months.

We will build our model around `status` and `time`, although we can examine the impact of the other variables later.

```{r}
#| warning: false

library(survival)
library(dplyr)


dataset <- lung |>
  mutate(
    # lung uses 1 and 2 for outcomes, but we want 0 and 1 for modeling
    status = recode(status, '1' = 0, `2` = 1),
    # Recode sex values for easier interpretation
    sex = recode(sex, '1' = 'M', `2` = 'F')
  )

# Surv() method creates a survival object
# survfit() fits the object to the data
survival_model <- survfit(Surv(time, status) ~ 1, data = dataset)
str(survival_model)
```

We can probably infer some conclusions from these numbers, but let's visualize them and get a better understanding of our model's curve.

```{r}
#| warning: false

library(ggsurvfit)

# Different fitting method used for ease of plotting
survfit2(Surv(time, status) ~ 1, data = dataset) |> 
  # Distinct plotting function for survival models
  ggsurvfit() +
  labs(
    x = "Days",
    y = "Overall survival probability"
  ) + 
  # Helper function to see CI
  add_confidence_interval()

```

Let's ask another question: How does our model predict the likelihood of survival at biannual benchmarks? To answer this, we can use the `tbl_survfit()` function from the `gtsummary`package:

```{r}
#| warning: false

library(gtsummary)

survival_model |>
  tbl_survfit(
    # Calculate probability every 6 months for first 2 years
    times = c(182.5, 365, 547.5, 720),
    label_header = "**{time} Days** (95% CI)"
  )
```

Lastly, let's analyze how survival likelihoods vary across qualitative or quantitative variables. The `survdiff()` function demonstrates how survival performance differs along qualitative variable lines. For example, the following excerpt compares how men and women fare compared to predicted survival rates:

```{r}
#| warning: false

# Create new Surv object with sex as dividing line
survdiff(Surv(time, status) ~ sex, data = dataset)
```

This output demonstrates that, in a model that doesn't account for sex by default, women have fewer observed deaths than projected, and the opposite is true for men. Considering that *p = 0.001*, it appears statistically significant that female patients are at a survival advantage when dealing with lung cancer compared to men.

If we want to measure the impact of a quantitative value on survival, the best course of action is to create a Cox regression model. The `coxph` function, which is included in the `survival` package, allows the user to create a Cox model based on an existing `Surv` object. The following example examines the effects of age on survival rates relative to predictions:

```{r}
#| warning: false

# Create new Surv object with age as dividing line
fit_cox <- coxph(Surv(time, status) ~ age, data = dataset)
summary(fit_cox)
```

As this model's value for `exp(coef)`, which represents the odds ratio of patients at *n* years relative to *n-1* years (all other factors held constant) is 1.019, each additional year of age increases a patient's risk of death by 1.9%. Additionally, a 95% confidence interval of *(1.001, 1.037)* demonstrates a 95% confidence that the true hazard of an increase in age results in an increase in risk. Finally, a *p* value of 0.04 shows that these findings are statistically significant.

In totality, the `survival` package is an excellent option for modeling any dataset that focuses on the progressing hazard of a key event.

### Mixed-Effects Modeling (lmer, glmer)

### Non-Parametric Machine Learning Modeling (knn)

k-Nearest Neighbors (k-NN) is a non-parametric method that utilizes supervised machine learning to classify data entries based on similarity to other entries. The `class` package in R, which contains a variety of machine learning functions, includes a method for building k-NN models, and will serve as the basis for our example implementation.

Let us imagine that we want to be able to classify flower species from the `iris` dataset from base principles. We can train a k-NN model with three defined classes and test its accuracy. This will require the application of a few distinct aspects of machine learning, such as:

* **Input Normalization** - scaling numerical input values relative to the dataset's observed maximum and minimum values for a given variable on a (0,1) or (-1, 1) scale.
* **Train-Test Dataset Splits** - isolating a minority of entries in a dataset to exclude from training and use to test the accuracy of the final model.
* **Confusion Matrices** - tables that highlight inaccurate classification predictions that machine learning models make.

With these concepts defined, we can begin to build our model. First, we will normalize the dataset's numeric columns so they are machine learning ready.

```{r}
#| warning: false

library(class)
library(dplyr)

iris_norm <- iris |>
mutate(
    across(where(is.numeric), \(x) (x - min(x)) / (max(x) - min(x)))
)
```

Next, we will split the dataset into an 80/20 train/test ratio using the `sample()` method.

```{r}
#| warning: false

set.seed(123)

train_indices <- sample(seq_len(nrow(iris_norm)), size = 0.8 * nrow(iris_norm))

train_df <- iris_norm |> slice(train_indices)
test_df  <- iris_norm |> slice(-train_indices)
```

Next, we will build the k-NN model itself. There are 4 parameters we need to define within the `knn()` method:

* `train` - the data to be used for training
* `test` - the data to be used for testing
* `cl` - the classification variable to predict
* `k` - the number of "neighborhoods" for the model to classify entries into

```{r}
#| warning: false

predictions <- knn(
  train = train_df %>% select(Sepal.Length:Petal.Width),
  test  = test_df  %>% select(Sepal.Length:Petal.Width),
  cl    = train_df$Species,
  k     = 3
)
```

Lastly, we will evaluate the model by comparing its predictions to the test dataset's actual classifications. 

```{r}
#| warning: false

table(Actual = test_df$Species, Predicted = predictions)
```

As the table shows, the model only made one incorrect prediction, a versicolor flower that was predicted to be virginica. Testing performance may vary due to seeding variance other factors, but in this case the model performed quite well.

All in all, k-NN models are strong options for projects where you would like to predict classifications based on a wholistic data frame.

<!-- do randomforest if you have time after finishing everything else?-->

### Other Noteworthy Models