---
title: "Performing Exploratory Data Analysis"
---

Once you have acquired a dataset to study, the first step in the analytical process is exploratory data analysis, or EDA. EDA is the process of observing and summarizing the key characteristics of a dataset to uncover basic underlying trends before diving deeper into advanced analysis. The exact steps of EDA that fit a project best will vary based on the dataset you're studying, but this article will lay out some of the paths you can take during this step of your analytical process.

### Descriptive Statistics

Descriptive statistics are basic measurements of dimension and spread that can give the user an understanding of the expected ranges and centers of a dataset's variables. The `psych` package in R contains a function `describe()` which is quite effective for generating a litany of descriptive statistics, as shown in the example below:

```{r}
#| warning: false 

library(psych)
describe(iris)
```

### Univariate Analysis

It helps to have descriptive statistics of a dataset, but if you want to get a better look at a single variable you may want to perform univariate analysis and visualize it. Regardless of whether the variable in question is quantitative or categorical, multiple avenues of univariate analysis lay at your disposal.

#### Categorical Univariate Analysis

1. Frequency Tables (Absolute and Relative)

A simple method to display the number of appearances of each option within a chosen variable. Ex:

```{r}
#| warning: false 

# Absolute Frequency Table
table(iris$Species)
# Relative Frequency table
prop.table(table(iris$Species))
```

2. Pie Charts & Bar Charts

Two simple plots to visually describe the characteristics of quantitative variables. Bar charts are more modular and easier to create in `ggplot2` whereas pie charts are generally more visually appealing to general audiences. Ex:

```{r}
#| warning: false 
library(ggplot2)
# Bar Chart
ggplot(data = iris, aes(x = Species)) +
geom_bar(aes(fill = Species)) +
theme_minimal()

# Pie Chart
ggplot(data = iris, aes(x = "", fill = Species)) +
geom_bar(color = "white", width = 1) +
coord_polar(theta = "y", start = 0) +
theme_minimal() +
labs(x = NULL, y = NULL)
```

#### Quantitative Univariate Analysis

1. Histograms

Charts that show the distribution of a variable in fine detail, consolidating points into small "bins" on the X axis and tallying the number of observations that fall into each bin on the Y axis. Ex:

```{r}
#| warning: false 
# Histogram
library(ggplot2)
ggplot(data = iris, aes(x = Petal.Length)) +
geom_histogram() +
theme_minimal()
```

2. Box Plots

Charts that visualize the general spread of an object in the form of key characteristics, including:

    a. Expected Minimums/Maximums (outer ticks)
    b. Q1 and Q3 indicators (box width)
    c. Median (middle dash in box)
    d. Outliers (dots outside outer ticks)

```{r}
#| warning: false 
# Box Plot
library(ggplot2)
ggplot(data = iris, aes(x = Petal.Length)) +
geom_boxplot() +
theme_minimal()
```

#### Footnote

All the following charts can be used in multivariate analysis as well through the addition of another variable as the `fill` factor (except the pie chart). Let's use the histogram from earlier as example to highlight how this added layer of complexity can reveal more about a dataset's underlying trends:

```{r}
#| warning: false 
library(ggplot2)
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
geom_histogram() +
theme_minimal()
```

### Bivariate Analysis

#### Bivariate Analysis - 2 Categorical Variables

1. Contingency Tables

Tables that tally the number of entries in a dataset that have each possible value across two or more variables. For our example, we will use a contingency table on `mtcars`, another example dataset from R, to count the number of cars with x cylinders and y carburetors.

```{r}
#| warning: false

table(
    as.factor(mtcars$cyl),
    as.factor(mtcars$carb),
    dnn=c("Cylinders", "Carburetors"))
```

Note: although the variables `cyl` and `carb` are technically numerical, we can use `as.factor()` to cast them to a categorical form, which can be helpful for numerical variables that behave categorically (such as cylinders in a car engine)

We can add a third variable to this tabulation as well, which will create a table for each of its values. In this example, we use `am`, which denotes whether a car has automatic (0) or manual (1) transmission.

```{r}
#| warning: false

table(
    as.factor(mtcars$cyl),
    as.factor(mtcars$carb),
    as.factor(mtcars$am),
    dnn=c("Cylinders", "Carburetors", "isManual"))
```

If it is more valuable to have these values as a representation of a percentage of a group, we can make these tables proportional using the `prop.table()` method. However, the group used as reference for determining each cell's proportion depends on the value chosen for `margin`.

* `margin` = 1: Proportional across rows (row sums = 100%)
* `margin` = 2: Proportional across columns (column sums = 100%)
* `margin` = NA: Proportional across all values (sum of all cells = 100%)

Let's see this in action with `mtcars` one more time, making a table along each `margin` value.

```{r}
#| warning: false

#Proportional Across Rows
prop.table(
    table(
        as.factor(mtcars$cyl),
        as.factor(mtcars$carb),
        dnn=c("Cylinders", "Carburetors")),
    margin = 1
)
#Proportional Across Columns
prop.table(
    table(
        as.factor(mtcars$cyl),
        as.factor(mtcars$carb),
        dnn=c("Cylinders", "Carburetors")),
    margin = 2
)
#Proportional Across All Values
prop.table(
    table(
        as.factor(mtcars$cyl),
        as.factor(mtcars$carb),
        dnn=c("Cylinders", "Carburetors"))
)
```

These proportional values make it very easy to begin to analyze the behavior of these categorical variables. For example, in `mtcars`:

* All 4-cylinder cars have only 1 or 2 carburetors.
* 71% of cars with 1 carburetor only have 4 cylinders.
* The two most common types of cars by cylinder and carburetor count are 4cyl-2carb and 8cyl-4carb, at 18.75% total proportion each.

2. Chi-Square Test of Independence

If you want to test the relationship between two categorical variables, you can perform a Chi-Square test to evaluate their independence. For example, let's use the `HairEyeColor` dataset and test if hair and eye color are related.

```{r}
#| warning: false
library(tidyr)
#Initial dataset contains tallies, revert to observations with uncount()
data <- as.data.frame(HairEyeColor) |>
  uncount(Freq)

#Data must be in table format for test
hair_eye_table <- table(data$Hair, data$Eye)

results <- chisq.test(hair_eye_table)
print(results)
```

Each of the output values from the test have their own importance, but what we care about most is the p-value, which can be interpreted as such:

> If there were no relationship between hair color and eye color in the overall population, the probability that their relationship would randomly appear as it does in this sample (p) is 2.2 * 10^-16^. Since p < 0.05, we can state that the result is statistically significant and reject the null hypothesis, H~0~, that there is no relationship between hair and eye color.

If you would like to learn more about performing chi-square tests in R, we recommend [this article](https://www.datacamp.com/tutorial/chi-square-test-r) for further reading.

#### Multivariate Analysis - 2 or More Quantitative Variables

1. Scatter Plot

Scatter plots are the simplest form of quantitative multivariate analysis, and can be transformed to add detail in many ways. In `ggplot`, attributes such as `size` and `color` can be assigned to a dataset's variables to express information in a digestible manner.

Below are two examples of themes within `iris` and `mtcars` visualized using scatter plots.

```{r}
#| warning: false
library(ggplot2)
ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length,
size = Petal.Length, color = Species)) +
geom_point() +
labs(
    title = "Sepal and Petal Size by Species",
    subtitle = "Data = iris, n = 150"
) +
theme_minimal()
```

```{r}
#| warning: false
library(ggplot2)
ggplot(data = mtcars, aes(x = hp, y = mpg,
size = cyl, color = as.factor(vs))) +
geom_point() +
labs(
    title = "Scatter Plot of MPG by horsepower",
    subtitle = "Data = mtcars, n = 32",
    color = "EngineIsStraight",
    size = "Cylinders",
    x = "Horsepower",
    y = "Miles Per Gallon (MPG)"
) +
theme_minimal()
```

2. Density Plot

Density plots are the two-dimensional equivalent of histograms, with counts expressed on a color scale since the Y-axis is no longer available. If you make a scatter plot of your dataset, but there are too many points to discern any valuable information, there are two best options: Reduce the `alpha` value, which controls opacity, in the `aes` field, or replace the scatter plot with a density plot.

The following example shows the strength of making a density plot when analyzing a large dataset, such as `diamonds`.

```{r}
#| warning: false
library(ggplot2)
data(diamonds)
diamonds |>
  ggplot(aes(x = carat, y = price)) + 
  geom_hex(bins = 70) +
  # Uses transform = "sqrt" for a gentler compression curve
  scale_fill_viridis_c(transform = "sqrt") +
  labs(
    title = "Heatmap of Diamond Price by Carat",
    subtitle = "Data = diamonds, n = 53940",
    x = "Carats",
    y = "Price (USD)"
  ) +
  theme_minimal()
```

3. Line Plot

For ordinal data, especially time-based variables, line plots can visualize trends in a highly digestible manner.

The following example uses the `nycflights13` package for acquisition and the `lubridate` package, an all-in-one tool for date and time manipulation, for transformation.

```{r}
#| warning: false
library(nycflights13)
library(ggplot2)
library(lubridate)
library(dplyr)

flights |>
  # Floor the date to the beginning of the week
  mutate(date = floor_date(time_hour, unit = "week")) |>
  # Exclude partial weeks
  filter(date > as_date("2012-12-30") & date < as_date("2013-12-29")) |>
  group_by(date) |>
  summarize(n_flights = n()) |>
  ggplot(aes(x = date, y = n_flights)) +
  geom_line() +
  # Reformat X-axis for readability
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 months") +
  labs(
    title = "2013 Total Flights per Week",
    subtitle = "Incomplete weeks omitted for visual clarity",
    y = "count"
  ) +
  theme_minimal()
```

4. Correlograms

Correlograms are plots that show the correlation value between all quantitative variables in a dataset. They can easily present a variety of leads that you could investigate as you build your hypothesis. Ex:

```{r}
#| warning: false

library(GGally)
library(ggplot2)

ggcorr(mtcars, 
       method = c("everything", "pearson"),
       label = TRUE,
       label_size = 3,
       label_round = 2 
       ) +
  labs(title = "mtcars Variable Correlogram") +
  theme_minimal()
```

This plot immediately presents a variety of leads, such as:

* As cylinders, displacement, horsepower, and weight increase, MPG tends to decrease.
* Manual cars have a strong positive correlation with gear count
* Carburetor count has a strong positive correlation with horsepower

### Conclusion

When performing EDA during project development, you likely won't need to use all these methods. There is a skill to understanding which process best fits your dataset, and that will come with time. When preparing to perform EDA, ask yourself:

* What variables appear most important? Are they categorical or quantitative?
* What variable relationships will define my study? How can I frame them initially?
* What findings from EDA would provide valuable context to my audience and are worth using in the final report?

If you can come up with well-thought-out answers to questions like these, EDA will become much easier.